---
title: "Data Cleaning"
author: "Andrew Edwards, Nick Oakley, and Lauren Pagel"
date: "2024-11-05"
output: html_document
---

## Package Installation and Initialization

In this portion we use `pacman` to install the necessary packages and clear the global environment.

```{r setup,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE, out.width = "100%")
#package initialization
rm(list = ls()) # clear global environment
graphics.off() # close all graphics
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(DataExplorer, # For explorer and cleaning data by visualizing missing values.
               httr, # For downloading the data from the UCI
               tidyverse, # For data manipulation
               corrplot, # for heatmaps
               DataExplorer, #for exploring data
               skimr, # for nice summary of data
               readxl, #to read Excel files
               networkD3 # to create a Sankey Plot
               ) # For creating dummy variables of categorical variables automatically.)
```

## Load data

### Loading BA Students

This is the list of students under consideration.  Here is the logic for this data pull:


```{r}
if(require(readxl)==FALSE) install.packages("readxl")

# Read the Excel file
BA_Students <- readxl::read_excel("BA Major Students.xlsx")

# Display the first few rows of the data
head(BA_Students)

# Print the number of rows and unique students
paste("The number of rows in the data is ", nrow(BA_Students))
paste("The number of unique students is ", length(unique(BA_Students$`Student ID`)))

# Check if the directory exists, if not, create it
if (!dir.exists("data-clean")) {
  dir.create("data-clean")
}

# Write the data to a CSV file
write.csv(BA_Students, "data-clean/BA_Students.csv")
```
### Loading BA Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
BA_Majors=read_excel("BA Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
BA_Majors=BA_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
#Create a separate column for each major.  Unit of analysis here is student by Term Code
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`BA Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(BA_Majors)
paste("The number of rows is ",nrow(BA_Majors))
paste("The number of unique students is ",length(unique((BA_Majors$`Student ID`))))

write.csv(BA_Majors,"data-clean/BA_Majors.csv")
```
### Loading Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
BA_Grades=read_excel("BA Major Students - Grades -updated.xlsx")
# Fill in the blank cells created from the format of the data
BA_Grades = BA_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(BA_Grades)
tail(BA_Grades)
paste("The number of unique students is ",length(unique(BA_Grades$`Student ID`)))


write.csv(BA_Grades,"data-clean/BA_Grades.csv")
```
### Loading Attributes

Below I load in the Attributes.  There are 134 students with missing cumulative GPA.  

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
BA_Attributes=read_excel("BA Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
BA_Attributes = BA_Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(BA_Attributes$`Student ID`)))

head(BA_Attributes)
test=BA_Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(BA_Attributes,"data-clean/BA_Attributes.csv")
```
```{r}
ba_major_crhr = read_excel("BA Major Students - Credit Hours.xlsx")

ba_major_crhr = ba_major_crhr %>% fill(c(`Student ID`),.direction="down")

paste("The number of unique students is ",length(unique(ba_major_crhr$`Student ID`)))

head(ba_major_crhr)
```
```{r}
ba_major_degrees_awarded = read_excel('BA Major Students - Degrees Awarded.xlsx')

ba_major_degrees_awarded = ba_major_degrees_awarded %>% 
  fill(c(`Student ID`:`Term Code`: 
           `Degree Type`: `Degree`),.direction="down")
head(ba_major_degrees_awarded)
```




```{r, echo=FALSE}
## join ba_major_final tables together 
ba_major_final = dplyr::left_join(BA_Majors, BA_Attributes, by = "Student ID")

ba_major_final <- dplyr::left_join(ba_major_final, BA_Grades, by = "Student ID")


```

# IS Student Cleaning


### Loading IS Students

This is the list of students under consideration.  Here is the logic for this data pull:



```{r}
IS_Students=read_excel("IS Major Students.xlsx")
head(IS_Students)
paste("The number of rows in the data is ",nrow(IS_Students))
paste("The number of unique students is ",length(unique(IS_Students$`Student ID`)))
write.csv(IS_Students,"data-clean/IS_Students.csv")

```

### Loading IS Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
IS_Majors=read_excel("IS Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
IS_Majors=IS_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
colnames(IS_Majors)
#Create a separate column for each major.  Unit of analysis here is student by Term Code
IS_Majors=IS_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
IS_Majors=IS_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(IS_Majors)
paste("The number of rows is ",nrow(IS_Majors))
paste("The number of unique students is ",length(unique((IS_Majors$`Student ID`))))

write.csv(IS_Majors,"data-clean/IS_Majors.csv")
```

### Loading Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
IS_Grades=read_excel("IS Major Students - Grades -updated.xlsx")
# Fill in the blank cells created from the format of the data
IS_Grades = IS_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(IS_Grades)
tail(IS_Grades)
paste("The number of unique students is ",length(unique(IS_Grades$`Student ID`)))


write.csv(IS_Grades,"data-clean/Grades.csv")
```

### Loading Attributes

Below I load in the Attributes.    

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
IS_Attributes=read_excel("IS Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
IS_Attributes = IS_Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(IS_Attributes$`Student ID`)))

head(IS_Attributes)
test=IS_Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(IS_Attributes,"data-clean/Attributes.csv")
```
```{r}
is_major_crhr = read_excel("IS Major Students - Credit Hours.xlsx")

is_major_crhr = is_major_crhr %>% fill(c(`Student ID`),.direction="down")

paste("The number of unique students is ",length(unique(is_major_crhr$`Student ID`)))

head(is_major_crhr)
```


```{r, echo=FALSE}
## join ia_major_final tables together 
is_major_final = dplyr::left_join(IS_Majors, IS_Attributes, by = "Student ID")

is_major_final <- dplyr::left_join(is_major_final, IS_Grades, by = "Student ID")




```

```{r, echo=FALSE}
## Combine BA and IS Data Together
final_df <- dplyr::bind_rows(ba_major_final, is_major_final)
head(final_df)
```

