---
title: "Data Cleaning"
author: "Andrew Edwards, Nick Oakley, and Lauren Pagel"
date: "2024-11-05"
output: html_document
---

## Package Installation and Initialization

In this portion we use `pacman` to install the necessary packages and clear the global environment.

```{r setup,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE, out.width = "100%")
#package initialization
rm(list = ls()) # clear global environment
graphics.off() # close all graphics
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(DataExplorer, # For explorer and cleaning data by visualizing missing values.
               httr, # For downloading the data from the UCI
               tidyverse, # For data manipulation
               corrplot, # for heatmaps
               DataExplorer, #for exploring data
               skimr, # for nice summary of data
               readxl, #to read Excel files
               networkD3 # to create a Sankey Plot
               ) # For creating dummy variables of categorical variables automatically.)
```

## Load data

### Loading BA Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
BA_Majors=read_excel("BA Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
BA_Majors=BA_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
#Create a separate column for each major.  Unit of analysis here is student by Term Code
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`BA Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(BA_Majors)
paste("The number of rows is ",nrow(BA_Majors))
paste("The number of unique students is ",length(unique((BA_Majors$`Student ID`))))

BA_Majors$`Major 4` = NULL
BA_Majors$`Major 5` = NULL
BA_Majors$`Major 6` = NULL

head(BA_Majors)



write.csv(BA_Majors,"data-clean/BA_Majors.csv")
```
### Loading Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
BA_Grades=read_excel("BA Major Students - Grades -updated.xlsx")
# Fill in the blank cells created from the format of the data
BA_Grades = BA_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(BA_Grades)
tail(BA_Grades)
paste("The number of unique students is ",length(unique(BA_Grades$`Student ID`)))


write.csv(BA_Grades,"data-clean/BA_Grades.csv")
```
### Loading Attributes

Below I load in the Attributes.  There are 134 students with missing cumulative GPA.  

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
BA_Attributes=read_excel("BA Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
BA_Attributes = BA_Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(BA_Attributes$`Student ID`)))

head(BA_Attributes)
test=BA_Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(BA_Attributes,"data-clean/BA_Attributes.csv")
```
```{r}
ba_major_crhr = read_excel("BA Major Students - Credit Hours.xlsx")

ba_major_crhr = ba_major_crhr %>% fill(c(`Student ID`),.direction="down")

paste("The number of unique students is ",length(unique(ba_major_crhr$`Student ID`)))

head(ba_major_crhr)
write.csv(ba_major_crhr, "data-clean/BA_CreditHours.csv")
```
```{r}
ba_major_degrees_awarded = read_excel('BA Major Students - Degrees Awarded.xlsx')

ba_major_degrees_awarded = ba_major_degrees_awarded %>% 
  fill(c(`Student ID`:`Term Code`: 
           `Degree Type`: `Degree`),.direction="down")
head(ba_major_degrees_awarded)

# Step 1: Create a separate column for each major (unit of analysis is student by Term Code)
ba_major_degrees_awarded <- ba_major_degrees_awarded %>%  
  group_by(`Student ID`, `Term Code`) %>%  
  mutate(Major_Number = paste("Major", row_number())) %>%  
  pivot_wider(names_from = Major_Number, values_from = `Major`) 

# Step 2: Ensure each student has only one row per term with all their majors side by side
# We'll use mutate to keep all columns, including Semesters to Degree Average
ba_major_degrees_awarded <- ba_major_degrees_awarded %>%
  group_by(`Student ID`, `Term Code`) %>%
  mutate(across(starts_with("Major"), ~ first(na.omit(.)), .names = "{.col}")) %>%
  ungroup()

# Step 3: Set the Cohort Term to the smallest observed Term Code for each student
ba_major_degrees_awarded <- ba_major_degrees_awarded %>%  
  group_by(`Student ID`) %>%  # Group by Student ID to find the minimum Term Code per student
  mutate(`Cohort Term` = min(`Term Code`)) %>%  # Set Cohort Term to the smallest Term Code for each student
  ungroup()  # Ungroup to return to a regular data frame


ba_major_degrees_awarded$`Student Division Code` = NULL
ba_major_degrees_awarded$`Student Department Code` = NULL
ba_major_degrees_awarded$`AW - Major Count` = NULL
ba_major_degrees_awarded$`Cohort Term` = NULL

# View the result
head(ba_major_degrees_awarded)

value_counts <- table(ba_major_degrees_awarded$`Semesters to Degree Average`)

# Print the value counts
print(value_counts)

write.csv(ba_major_degrees_awarded, "data-clean/BADegrees.csv")
```




# IS Student Cleaning


### Loading IS Students

This is the list of students under consideration.  Here is the logic for this data pull:


### Loading IS Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
IS_Majors=read_excel("IS Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
IS_Majors=IS_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
colnames(IS_Majors)
#Create a separate column for each major.  Unit of analysis here is student by Term Code
IS_Majors=IS_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
IS_Majors=IS_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(IS_Majors)
paste("The number of rows is ",nrow(IS_Majors))
paste("The number of unique students is ",length(unique((IS_Majors$`Student ID`))))

IS_Majors$`Major 4` = NULL
IS_Majors$`Major 5` = NULL
IS_Majors$`Major 6` = NULL

head(IS_Majors)

write.csv(IS_Majors,"data-clean/IS_Majors.csv")
```

### Loading Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
IS_Grades=read_excel("IS Major Students - Grades -updated.xlsx")
# Fill in the blank cells created from the format of the data
IS_Grades = IS_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(IS_Grades)
tail(IS_Grades)
paste("The number of unique students is ",length(unique(IS_Grades$`Student ID`)))


write.csv(IS_Grades,"data-clean/IS_Grades.csv")
```

### Loading Attributes

Below I load in the Attributes.    

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
IS_Attributes=read_excel("IS Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
IS_Attributes = IS_Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(IS_Attributes$`Student ID`)))

head(IS_Attributes)
test=IS_Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(IS_Attributes,"data-clean/IS_Attributes.csv")
```
```{r}
is_major_crhr = read_excel("IS Major Students - Credit Hours.xlsx")

is_major_crhr = is_major_crhr %>% fill(c(`Student ID`),.direction="down")

paste("The number of unique students is ",length(unique(is_major_crhr$`Student ID`)))

head(is_major_crhr)
write.csv(is_major_crhr, "data-clean/IS_CreditHours.csv")
```

I will now clean the major degrees awared file for IS Majors. One field that is of particular interest to me is how many semesters it took a student to graduate, which can be displayed by the average of that field below. We found that, overwhelmingly, it took students 8 semester of coursework at Miami before they graduated. This provides our team with the valuable insight that graduation is likely better shown by the number of semesters of enrollment as opposed to number of credit hours completed.
```{r}
is_major_degrees_awarded = read_excel('IS Major Students - Degrees Awarded (1).xlsx')

is_major_degrees_awarded = is_major_degrees_awarded %>% 
  fill(c(`Student ID`:`Term Code`: 
           `Degree Type`: `Degree`),.direction="down")
head(is_major_degrees_awarded)

# Step 1: Create a separate column for each major (unit of analysis is student by Term Code)
is_major_degrees_awarded <- is_major_degrees_awarded %>%  
  group_by(`Student ID`, `Term Code`) %>%  
  mutate(Major_Number = paste("Major", row_number())) %>%  
  pivot_wider(names_from = Major_Number, values_from = `Major`) 

# Step 2: Ensure each student has only one row per term with all their majors side by side
# We'll use mutate to keep all columns, including Semesters to Degree Average
is_major_degrees_awarded <- is_major_degrees_awarded %>%
  group_by(`Student ID`, `Term Code`) %>%
  mutate(across(starts_with("Major"), ~ first(na.omit(.)), .names = "{.col}")) %>%
  ungroup()

# Step 3: Set the Cohort Term to the smallest observed Term Code for each student
is_major_degrees_awarded <- is_major_degrees_awarded %>%  
  group_by(`Student ID`) %>%  # Group by Student ID to find the minimum Term Code per student
  mutate(`Cohort Term` = min(`Term Code`)) %>%  # Set Cohort Term to the smallest Term Code for each student
  ungroup()  # Ungroup to return to a regular data frame

is_major_degrees_awarded <- is_major_degrees_awarded[-211, ]

is_major_degrees_awarded$`Student Division Code` = NULL
is_major_degrees_awarded$`Student Department Code` = NULL
is_major_degrees_awarded$`AW - Major Count` = NULL
is_major_degrees_awarded$`Cohort Term` = NULL

# View the result
head(is_major_degrees_awarded)

value_counts <- table(is_major_degrees_awarded$`Semesters to Degree Average`)

# Print the value counts
print(value_counts)

write.csv(is_major_degrees_awarded, "data-clean/ISDegrees.csv")
```
```{r, echo=FALSE}
## Merge BA Datasets together

## BA Degrees
BA_Degrees = read.csv('data-clean/BADegrees.csv')
BA_Degrees <- BA_Degrees |>
  dplyr::select(-c('X', 'Major.1', 'Major.2', 'Major.3', 'Degree.Type', 'Degree'))
BA_Degrees <- BA_Degrees[complete.cases(BA_Degrees), ]
BA_Degrees$Graduated <- ifelse(!is.na(BA_Degrees$`Semesters.to.Degree.Average`), 1, 0)
BA_Degrees$`Student.ID` <- as.integer(BA_Degrees$`Student.ID`)

```

```{r}
## BA Majors
BA_Majors = read.csv('data-clean/BA_Majors.csv')
head(BA_Majors)
BA_Majors <- BA_Majors |>
  dplyr::select(-c('X', 'Enrolled.Student.Count'))

BA_Majors <- BA_Majors |>
  dplyr::mutate(
    `Major.1` = tidyr::replace_na(`Major.1`, 'None'),
    `Major.2` = tidyr::replace_na(`Major.2`, 'None'),
    `Major.3` = tidyr::replace_na(`Major.3`, 'None')
  )

head(BA_Majors)

```
```{r}
## merge degrees and majors together
BA <- dplyr::left_join(BA_Majors, BA_Degrees, by = 'Student.ID')

## drop columns
BA <- BA |>
  dplyr::select(-c('Term.Code.y', 'Semesters.to.Degree.Average'))
## rename columm
BA <- BA |>
  dplyr::rename(`Term.Code` = `Term.Code.x`)

## Fill NA values in graduated. Graduated will equal 1 if anywhere in the dataset for a student id they have graduated. It will be 0 if not.

# Determine if each student has graduated
graduated_status <- BA |>
  dplyr::group_by(`Student.ID`) |>
  dplyr::summarise(Graduated = max(Graduated, na.rm = TRUE)) |>
  dplyr::mutate(Graduated = ifelse(Graduated == -Inf, NA_real_, Graduated))

# Fill NA values in 'Graduated' based on Student ID
BA <- BA |>
  dplyr::left_join(graduated_status, by = "Student.ID", suffix = c("", ".y")) |>
  dplyr::mutate(Graduated = ifelse(is.na(Graduated.y), 0, Graduated.y)) |>
  dplyr::select(-Graduated.y)

head(BA)
```
```{r, echo=FALSE}
BA_Credit_Hours = read.csv('data-clean/BA_CreditHours.csv')
# drop columns
BA_Credit_Hours <- BA_Credit_Hours |>
  dplyr::select(-c('X', 'Term.UG.MU.Earned.CrHr', 'Term.UG.Overall.Earned.CrHr..MU.and.Non.MU.'))
# rename columns
BA_Credit_Hours <- BA_Credit_Hours |>
  dplyr::rename(`Cum.Credit.Hrs` = `Cum.UG.Overall.Earned.CrHr..MU.and.Non.MU.`)
# fill na
BA_Credit_Hours <- BA_Credit_Hours |>
  dplyr::mutate(`Cum.Credit.Hrs` = tidyr::replace_na(`Cum.Credit.Hrs`, 0))
head(BA_Credit_Hours)
```
```{r}
## merge credit hours with BA dataset
BA <- dplyr::left_join(BA, BA_Credit_Hours, by = c("Student.ID", "Term.Code"))
BA <- BA |>
  dplyr::arrange(`Student.ID`)

## filter out error
BA <- BA |>
  dplyr::filter(`Major.1` != "None")

## replace NA Credit Hrs with 0
BA <- BA |>
  dplyr::mutate_all(~ tidyr::replace_na(., 0))
head(BA)
```
```{r, echo=FALSE}
BA_Attributes = read.csv('data-clean/BA_Attributes.csv')
# drop columns
BA_Attributes <- BA_Attributes |>
  dplyr::select(-c('X', 'Gender', 'First.Generation.Indicator', 'Term.UG.Crs.GPA'))

## fill NA values
BA_Attributes <- BA_Attributes |>
  dplyr::mutate(`Cum.UG.Crs.GPA` = ifelse(is.na(`Cum.UG.Crs.GPA`), "NA", `Cum.UG.Crs.GPA`))
head(BA_Attributes)
```

```{r}
## merge attributes and BA data set together
BA <- dplyr::left_join(BA, BA_Attributes, by = c("Student.ID", "Term.Code"))
BA <- BA |>
  dplyr::arrange(`Student.ID`)
head(BA)
write.csv(BA, "data-clean/BA_Final.csv")
```



