---
title: "BA Student Movement Data Cleaning"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "Andrew Edwards, Nick Oakley, Lauren Pagel"
format: 
  html:
    toc: true
    toc-location: left
    toc-expand: true
    code-tools: true 
    code-fold: true
    code-overflow: scroll
    self-contained: true
    #bibliography: winerefs.bib
---


## Package Installation and Initialization

In this portion we use `pacman` to install the necessary packages and clear the global environment.

```{r setup,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE, out.width = "100%")
#package initialization
rm(list = ls()) # clear global environment
graphics.off() # close all graphics
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(DataExplorer, # For explorer and cleaning data by visualizing missing values.
               httr, # For downloading the data from the UCI
               tidyverse, # For data manipulation
               corrplot, # for heatmaps
               DataExplorer, #for exploring data
               skimr, # for nice summary of data
               readxl, #to read Excel files
               networkD3 # to create a Sankey Plot
               ) # For creating dummy variables of categorical variables automatically.)
```

## Load data

### Loading BA Students

This is the list of students under consideration.  Here is the logic for this data pull:


```{r}
if(require(readxl)==FALSE) install.packages("readxl")

# Read the Excel file
BA_Students <- readxl::read_excel("BA Major Students.xlsx")

# Display the first few rows of the data
head(BA_Students)

# Print the number of rows and unique students
paste("The number of rows in the data is ", nrow(BA_Students))
paste("The number of unique students is ", length(unique(BA_Students$`Student ID`)))

# Check if the directory exists, if not, create it
if (!dir.exists("data-clean")) {
  dir.create("data-clean")
}

# Write the data to a CSV file
write.csv(BA_Students, "data-clean/BA_Students.csv")
```


### Loading BA Majors

Below I read in the data and fill in the missing values with the correct values.  The level of analysis is Student ID X Term Code.  Each student has a row for each term they have been enrolled.  Some students have one row, others have eight.

Finally, I make sure that each student has the Cohort Term that is equal to the earliest date on record.

```{r}
# Read data
BA_Majors=read_excel("BA Major Students - Majors.xlsx")
# Fill in the blank cells created from the format of the data
BA_Majors=BA_Majors %>% 
  fill(c(`Student ID`:`Term Code`),.direction="down")
#Create a separate column for each major.  Unit of analysis here is student by Term Code
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`,`Term Code`) %>% 
  mutate(Major_Number=paste("Major",row_number())) %>% 
  pivot_wider(names_from=Major_Number, values_from=`BA Major`) 

# Set the Cohort Term to the smallest observed value for a given student. 
BA_Majors=BA_Majors %>% 
  group_by(`Student ID`) %>% 
  mutate(`Cohort Term`=min(`Cohort Term`)) 

head(BA_Majors)
paste("The number of rows is ",nrow(BA_Majors))
paste("The number of unique students is ",length(unique((BA_Majors$`Student ID`))))

write.csv(BA_Majors,"data-clean/BA_Majors.csv")
```

### Loading Grades

Loaded student grades, filled in missing cells according to the data structure.  There are only 694 students.  This is possibly because the freshmen do not yet have grades. I checked this, and there are 140 freshmen who have no grades.  694+140=834.  We have 823 unique students in the file.  It is possible that 834+823=11 students in cohort 202410 have grades, possibly due to dual enrollment or summer classes prior to the Fall semester.

```{r}
# Read data
BA_Grades=read_excel("BA Major Students - Grades.xlsx")
# Fill in the blank cells created from the format of the data
BA_Grades = BA_Grades %>% 
  fill(c(`Student ID`),.direction="down")

head(BA_Grades)
tail(BA_Grades)
paste("The number of unique students is ",length(unique(BA_Grades$`Student ID`)))
num_fresh=BA_Majors %>%
  filter(`Cohort Term`=="202410")%>%
  nrow()
paste("The number of students starting in term 201410 is ", num_fresh)

write.csv(BA_Grades,"data-clean/BA_Grades.csv")
```

### Loading Attributes

Below I load in the Attributes.  There are 134 students with missing cumulative GPA.  

The level of analysis for the final data frame is Student ID X Term Code.

```{r}
# Read data
BA_Attributes=read_excel("BA Major Students - Attributes and GPA.xlsx")
# Fill in the blank cells created from the format of the data
BA_Attributes = BA_Attributes %>% 
  fill(c(`Student ID`:`First Generation Indicator`),.direction="down")

paste("The number of unique students is ",length(unique(BA_Attributes$`Student ID`)))

head(BA_Attributes)
test=BA_Attributes %>% 
  filter(is.na('Term UG Crs GPA'))

paste("The number of students with no grades ",length(unique(test$`Student ID`)))
write.csv(BA_Attributes,"data-clean/BA_Attributes.csv")
```

